\subsection*{Attachment A: Digital Implementation Archive}

A complete digital archive of the implementation (\texttt{shai.zip}) has been
uploaded to the university's official academic information system. This archive
serves as a permanent, immutable record of the project at the time of
submission.

\noindent \textbf{Archive Structure:}
\begin{itemize}
    \item \texttt{shAI-l/shAI\_ostap4ello/src/} -- Python source code of the CLI assistant and its components.
    \item \texttt{shAI-l/README.md} -- Setup instructions, prerequisites, and steps to replicate the experiments.
    \item \texttt{shAI-l/test/} -- Empirical tests and test collections for the software.
    \item \texttt{shAI-l/test/README.md} -- Test framework instructions.
\end{itemize}

\vspace{0.8cm}
\subsection*{Appendix B: Live Git Repository}

For ease of code review, access to the full development history (commit logs),
and potential future updates, the project is also hosted publicly on GitHub.

\vspace{0.3cm}
\noindent \textbf{Repository URL:} \\
\url{https://github.com/Ostap4ello/shAI-l.git}

\vspace{0.3cm}
\noindent \textit{Note: The source code provided in the official university
archive is identical to the repository state at the time of submission}
\textit{(Git commit hash: \texttt{a44f4a})}.

\vspace{0.8cm}
\subsection*{Appendix C: UML diagrams source code}

A archive (\texttt{diagrams.zip}) containing UML diagrams used in the thesis is
provided with the thesis.

\noindent \textbf{Archive Structure:}
\begin{itemize}
    \item \texttt{usecases.puml} --- Use-case diagram for ShAI.
    \item \texttt{component.puml} --- Component diagram for ShAI.
\end{itemize}
