ShAI is a procedural, CLI-driven application that combines local LLM access, FAISS-based retrieval,
and Ollama container management. The architecture is organized by package and file, with each module
owning a small, explicit responsibility.
\label{a-shai-architecture}

\subsection{Package tree}

\begin{verbatim}
shAI_ostap4ello/src/
  __main__.py
  config/
    __init__.py
    config.py
  db/
    __main__.py
    db.py
    utils/
      config.py
      documents.py
      faiss_utils.py
  interpreter/
    interpreter.py
  llm/
    __main__.py
    llm.py
  utils/
    __main__.py
    adapter.py
    fetch_man.py
    prompt.py
  workflows/
    __init__.py
    rag.py
    classifier_bash.py
\end{verbatim}

\subsection{Top-level entry point}

\paragraph{\texttt{src/\_\_main\_\_.py}}
Loads configuration, prepares the main CLI parser, and dispatches to the selected subcommand.
It also exposes the \texttt{init} flow, which creates the database folder, starts Ollama if needed,
pulls configured models, and removes the temporary container.

\subsection{Configuration package}

\paragraph{\texttt{src/config/config.py}}
Implements centralized configuration loading and default propagation. It reads
\texttt{\textasciitilde/.config/shai/config.conf}, resolves defaults such as the database path and
Ollama URL, and updates module-level values so the rest of the application sees one consistent
runtime configuration.

\paragraph{\texttt{src/config/\_\_init\_\_.py}}
Re-exports the public configuration helpers for convenient imports from other packages.

\subsection{LLM package}

\paragraph{\texttt{src/llm/llm.py}}
Contains the core OpenAI-compatible Ollama client helpers. It provides client construction,
non-streaming generation, streaming generation, single-string embeddings, and batched embeddings.

\paragraph{\texttt{src/llm/\_\_main\_\_.py}}
Exposes the \texttt{llm generate} and \texttt{llm embed} commands. It is the user-facing CLI wrapper
around \texttt{llm.py}.

\paragraph{\texttt{src/llm/\_\_init\_\_.py}}
Exports the reusable generation and embedding helpers for other modules.

\subsection{Database package}

\paragraph{\texttt{src/db/\_\_main\_\_.py}}
Implements the CLI for document indexing and retrieval. It wires the build, search, and check
commands, resolves path options, and connects CLI flags to the database layer.

\paragraph{\texttt{src/db/db.py}}
Contains the database orchestration logic. It builds FAISS indexes, performs search, exposes metadata
inspection, and provides update/add/remove operations around the stored document metadata.

\paragraph{\texttt{src/db/utils/config.py}}
Defines the index metadata format and resolves on-disk index paths. It is responsible for the hidden
index directory layout and the persisted database configuration file.

\paragraph{\texttt{src/db/utils/documents.py}}
Discovers readable source documents and loads them either as full files or as section-sized chunks.
It also filters out hidden paths and unsupported extensions.

\paragraph{\texttt{src/db/utils/faiss\_utils.py}}
Handles the in-memory FAISS index creation and the persistence layer for the vector index and metadata.
It stores the index file and loads it back for search.

\subsection{Workflow package}

\paragraph{\texttt{src/workflows/rag.py}}
Builds the retrieval-augmented generation flow. It converts search results into readable context,
loads referenced file content, and asks the LLM to answer using the retrieved documents.

\paragraph{\texttt{src/workflows/classifier\_bash.py}}
Implements the bash-vs-natural-language classifier used by the interactive shell.

\paragraph{\texttt{src/workflows/\_\_init\_\_.py}}
Re-exports workflow helpers so higher-level modules can compose retrieval and classification logic
without importing internal files directly.

\subsection{Utilities package}

\paragraph{\texttt{src/utils/adapter.py}}
Provides the system integration layer. It runs the Ollama Docker helper script, starts and stops the
container, checks container status, waits for API readiness, and manages models through the Ollama
HTTP API.

\paragraph{\texttt{src/utils/\_\_main\_\_.py}}
Exposes utility commands such as container lifecycle management, model management, and man-page
conversion.

\paragraph{\texttt{src/utils/fetch\_man.py}}
Contains the man-page extraction and conversion pipeline used by the utility CLI.

\paragraph{\texttt{src/utils/prompt.py}}
Loads text prompts used by the workflow layer.

\paragraph{\texttt{src/utils/\_\_init\_\_.py}}
Re-exports the utility helpers for use across the application.

\subsection{Interactive shell package}

\paragraph{\texttt{src/interpreter/interpreter.py}}
Implements the interactive loop. It ensures Ollama is available, optionally starts it, and uses the
LLM classifier to route user input while the shell is running.

\subsection{Runtime flow}

\begin{enumerate}
    \item The user invokes \texttt{shai} or a module-specific command.
    \item The relevant \texttt{\_\_main\_\_.py} parser resolves arguments and configuration defaults.
    \item Database commands index or search documents through the FAISS-backed storage layer.
    \item Workflow commands combine retrieval and generation to produce a final answer.
    \item Utility commands manage Ollama and related local tooling.
    \item The shared configuration layer keeps the paths, models, and service endpoints aligned.
\end{enumerate}

\subsection{Design characteristics}

\begin{itemize}
    \item \textbf{Procedural style}: logic is organized as functions and modules rather than classes.
    \item \textbf{Local-first execution}: all model access goes through a local Ollama service.
    \item \textbf{Explicit layering}: each package owns one part of the stack, from CLI parsing to
    retrieval to generation.
    \item \textbf{Shared defaults}: configuration is centralized so every command sees the same runtime
    state.
\end{itemize}
