% TODO

This thesis presented an exploratory investigation into building a Linux
\gls{cli} assistant powered by local \glspl{llm}. The work evaluates multiple
architectural approaches, identifying which are practical for offline,
resource-constrained environments and which prove unsuitable for reliable
real-world use.

\subsection{Research Approach and Scope}

The work investigated three main architectural patterns for
\gls{llm}-based \gls{cli} assistance:
\begin{enumerate}
  \item \textbf{Natural Language Classification} --- Using
    \glspl{llm} to distinguish between shell commands and natural
    language queries
  \item \textbf{Agent-Based Reasoning} --- Enabling iterative
    reasoning, tool use, and decision-making for complex tasks
  \item \textbf{Retrieval-Augmented Generation (RAG)} --- Combining
    document retrieval with focused \gls{llm} summarization
\end{enumerate}

\subsection{Key Findings}

Through implementation and evaluation, significant practical insights emerged:

\paragraph{Classifier Component --- Not Viable}
LLM-based classification of commands vs.\ queries performs poorly with small local models
(0.5B-4B parameters). Achieving acceptable accuracy requires either: (a) unacceptable query
latencies (2-5+ seconds per input), or (b) extensive hardcoding of heuristics. Additionally,
seamless shell integration requires shell-specific implementations that are difficult to
maintain. This component remains a proof-of-concept only.

\paragraph{Agent-Based Approaches --- Unreliable}
Agentic systems that attempt multi-step reasoning, tool use, and state management consistently
fail with small local \glspl{llm}. Common failure modes include: reasoning chain breakdowns,
incorrect tool invocation, hallucination under uncertainty, and unacceptable latencies
(5-15 seconds per query). These approaches require significantly larger models (10B+ parameters)
or external services to function reliably.

\paragraph{RAG System --- Practical and Effective}
The Retrieval-Augmented Generation approach proved robust and practical. By constraining the
\gls{llm} task to extracting and summarizing information from retrieved documents, the system
achieves: reliable performance (minimal hallucination), acceptable latency (250-700ms per query),
and good user experience. This component forms the core working system.

\subsection{Implications for Local LLM Assistance}

The findings suggest that local \gls{llm} assistance on consumer hardware is best achieved
through \textbf{constrained, task-specific architectures} rather than general-purpose reasoning:

\begin{itemize}
  \item Limit task scope to information retrieval and summarization
  \item Avoid open-ended classification, reasoning, or tool-use tasks
  \item Provide focused context (via retrieval) to ground model outputs
  \item Accept trade-offs: helpful within constraints beats unreliable when pushed beyond them
\end{itemize}

\subsection{Practical Value}

ShAI provides practical offline assistance for \gls{cli} users, delivering fast,
help on user document datatbase without network dependency, latency, or privacy concerns. The system excels at: explaining commands, providing usage examples, and retrieving documentation.
The current implementation demonstrates that the \gls{rag} approach is viable and useful,
despite being less ambitious than original goals of general-purpose command-line reasoning.
This opens the door for further development, and, in focus of local \gls{llm}
assistance, suggests shift to embedding-based features rather and simpler
\gls{llm}-based applications rather than open-ended and agentic ones.

\subsection{Future Work and Extensions}

Future research can explore several avenues to build on these findings:
\begin{itemize}
  \item \textbf{Model Optimization:} Explore model distillation, quantization,
    or fine-tuning to further improve performance and reduce latency on local
    hardware.
  \item \textbf{User Interface Improvements:} Develop richer interfaces (e.g.,
    interactive query refinement, multi-turn conversations) while maintaining
    low latency and simplicity.
  \item \textbf{Broader Task Support:} Investigate additional constrained tasks
    that can be reliably handled by local models (e.g., code generation from
    templates, error message explanations).
  \item \textbf{Usage of RNNs:} Explore the actively researched area of
    \gls{rnn}-augmented \gls{llm} architectures.
  \item \textbf{Program-Aided Language Models (PAL) / Program-of-Thought (PoT):}
    Explore the novel approach of programmatic approach for \gls{llm} agentic
    workflows.
\end{itemize}

\subsection*{Contributions}

This work contributes to understanding practical \gls{llm} integration in terminal environments:

\begin{itemize}
  \item Empirical evidence of which architectures work with small local \glspl{llm}
  \item Identification of key failure modes in applying local models to structured tasks
  \item An open-source reference implementation of a working \gls{rag}-based assistant
  \item Guidance for future research: local model assistance is viable within constrained,
    well-scoped tasks, not as general-purpose reasoning
\end{itemize}

The thesis demonstrates that while ambitious visions of local
\gls{llm} assistants handling arbitrary tasks remain impractical,
specialized, constrained systems can provide real value to users with
various limitations. This work lays the groundwork for further
exploration \gls{llm}-enabled information processing and manipulation.
