
% \newpage
\section{Theoretical Background}\label{sec:theoretical-background}

This section provides the necessary theoretical background for understanding the
field of the thesis, concepts of technologies used, and the rationale behind the
implementation of the tools we will be using.
Its main goal is to provide the reader with sufficient information to understand
the rest of the thesis.

\input{includes/core/1-theoretical-background.tex}

%%%

\clearpage

\section{Analysis Of Existing Solutions And Technologies}

This section provides a comprehensive review of existing solutions, tools and
applications for chosen research topic.
During the review, the focus of the thesis has changed, due to already existing
solutions and conducted research in the field. The review describes shifts in
the focus of the thesis, and provides a rationale for the chosen focus.
Its main goal is to provide the reader with overview of the existing solutions
and technologies, and to justify the chosen focus of the thesis.

\input{includes/core/2-existing-solutions.tex}

%%%

\clearpage

\section{Design and Implementation}

This section provides a detailed description of the architectural design,
implementation details, and describes the considerations and decisions made
during the design and implementation of the solution.
Its main goal is to provide the reader with comprehensive overview of the design
and implementation of the solution, and experiments conducted during the design
and implementation process.

\input{includes/core/3-design-implementation.tex}

%%%

\clearpage

\section{Evaluation}

This section describes the evaluation proces for the application. For each
testing focus, it describes the chosen metrics, testing collection, and presents
the results of testing.
Its main goal is to provide the reader with comprehensive overview of tests and
their results, creating a background for the discussion of results and
conclusion on this thesis.

\input{includes/core/4-evaluation.tex}
