Fast development of technology in recent years, especially in computing power
and data storage, has enabled the \gls{ml} field to evolve rapidly.
This led us to the development of versatile \gls{ml} models, which are used in wide
range of applications.\ \glspl{llm} are one of the most popular types of \gls{ml} models
nowadays, which has shown great advancements and now has a lot of practical
applications.
\newline

The nature of the interaction with \gls{llm}'s makes us consider them as helpful
assistants in various tasks. The most appreciated feature of \glspl{llm} is their
ability to transfer from natural language instructions to executable commands.
\newline

One of common fields, where such feature can be utilized and increase the
\gls{qol} of users (from developers to regular \gls{os} users, as now Linux
distributions are becoming more user-friendly and popular), is the \gls{cli}. It
is to remain an essential part of modern development (and related stuff)
processes as it enables us to perform tasks quickly and efficiently.
\newline

This is what motivates us to create and evaluate a command line assistant, which
leverages \gls{llm} to break the barrier between \gls{natlang} instructions and
computer commands, yielding the need to learn specific syntax for each command
from the user.
\newline

The limitations of remote \glspl{llm} often present significant challenges in
real-world scenarios. Relying on a network connection introduces latency,
potential downtime, and dependency on external services.  Furthermore, data
privacy and security concerns frequently restrict the use of external \glspl{llm},
especially when dealing with sensitive information or proprietary tools.
Consider the scenario of a developer needing to quickly execute a complex
series of commands for debugging, or a system administrator requiring immediate
assistance in resolving a system issue – a network outage could be
devastating.
\newline

Therefore, this thesis aims to evaluate local \glspl{llm} as a viable solution for
\gls{cli} assistance. By running these models directly on the user's
machine, we eliminate the dependency on external services, reduce latency, and
address concerns regarding data privacy. This approach allows for a focused
investigation into the performance and usability of \glspl{llm} in a low-latency,
secure environment.\newline

The main goal of this thesis is to describe the process of design, development,
and to evaluate a Linux \gls{cli} assistant powered by different local \gls{llm} models
(differing in architecture, size, and capabilities and under different
constraints) to see how well they perform in various usage scenarios.
\newline

