
\subsection{Terminal and Command Line Interfaces and Applications}

This section describes the interfaces and programs that are relevant to our
solution. It defines two interfaces: \glspl{cli} and
\glspl{tui}. It also describes the programs that leverage
these interfaces: shells, \gls{cli} utilities, and \gls{tui} applications.

For the beginning, we need to understand what \glspl{cli} are, as well as the
utilities that are used within them.\\

The definition of the interface is adopted from Branden Hookway's book
\textit{Interface}, where he defines it as follows:
\begin{quote}
  \textit{\uv{In computing, an interface is a shared boundary across which
      two or more separate components of a computer system exchange information.
      The exchange can be between software, computer hardware, peripheral
  devices, humans, and combinations of these}}\cite{hookway2014}.
\end{quote}\\

\subsubsection{Terminal emulator, Terminal Application}

The thesis focuses on \glspl{tui}, which are a specific type of interface used
in computing environments. Here is how we define terminals, terminal emulators
and terminal applications.\\

\textbf{Terminal} is a \uv{\textit{an electronic or electromechanical hardware
    device that can be used for entering data into, and transcribing
data from, a computer or a computing system}}. In essence, the main
functionality of terminals and terminal emulators, which is to provide an
elementary interface for user to interact with the computer system.\\

\textbf{Terminal emulator/Terminal application} is a software application
that replicates the functionality of a traditional hardware terminal as a
software program. Here are some examples of terminal emulators:
\begin{itemize}
  \item iTerm2 (macOS)
  \item Windows Terminal (Windows)
  \item Alacritty (Cross-platform)
\end{itemize}

\textbf{Terminal applications} are software programs that run within terminal
emulators and utilize their capabilities to provide text-based user interfaces
--- either \glspl{cli} or \glspl{tui}. We will discuss it further in
the following sections\cite{man_man,linfo_terminal}.\\

% NOTE: this knowledge is not relevant for the solution. Delete later?
%
% In most cases, always, when the application is POSIX-compatible, we can define
% text streams, used by terminal application for
% communication.\cite{wiki_terminal}\cite{wiki_terminal_emulator} \\
%
% \begin{enumerate}
%   \item STDIN --- used for input fron user; initially as file descriptor 0.
%   \item STDOUT --- used for application output; initially as file
% descriptor 1.
%   \item STDIN --- used for error output; initially as file descriptor 2.
% \end{enumerate}
% \cite{open_group} \\
% POSIX on default communication streams for application: (hidden)\\
% \textit{\uv{
%     3.353 Standard Error
%
%     In the context of file descriptors (see 3.141 File Descriptor),
% file descriptor number 2.
%
%     In the context of standard I/O streams (see XSH 2.5 Standard
% I/O Streams), an output stream
%     usually intended to be used for diagnostic messages, and
% accessed using the global variable
%     stderr.
%
%     Note:
%     The file descriptor underlying stderr is initially 2, but it
% can be changed by freopen() to 0
%     or 1 (and implementations may have extensions that allow it to
% be changed to other numbers).
%     Therefore, writing to the standard error stream does not always
% produce output on the standard error
%     file descriptor.
%
%     3.354 Standard Input
%
%     In the context of file descriptors (see 3.141 File Descriptor),
% file descriptor number 0.
%
%     In the context of standard I/O streams (see XSH 2.5 Standard
% I/O Streams), an input stream usually
%     intended to be used for primary data input, and accessed using
% the global variable stdin.
%
%     Note:
%     The file descriptor underlying stdin is initially 0; this
% cannot change through the use of
%     interfaces defined in this standard, but implementations may
% have extensions that allow it to be
%     changed. Therefore, in conforming applications using
% extensions, reading from the standard input
%     stream does not always obtain input from the standard input
% file descriptor.
%
%     3.355 Standard Output
%
%     In the context of file descriptors (see 3.141 File Descriptor),
% file descriptor number 1.
%
%     In the context of standard I/O streams (see XSH 2.5 Standard
% I/O Streams), an output stream
%     usually intended to be used for primary data output, and
% accessed using the global variable
%     stdout.
%
%     Note: The file descriptor underlying stdout is initially 1, but
% it can be changed by freopen() to 0
%     (and implementations may have extensions that allow it to be
% changed to other numbers). Therefore,
%     writing to the standard output stream does not always produce
% output on the standard output file
%     descriptor.
% }}

\subsubsection{Shells and \gls{cli} Utilities}

The definition of \gls{cli} extends previous definition to the specific case of
command line interfaces, which are text-based interfaces between the user and
the computer system, where user inputs commands in the form of text, to
manipulate the system or perform specific tasks, and gets responses in the same
form.\\

\textbf{Shells} are special programs (with \gls{cli}, \gls{tui}, as well as
\gls{gui} interfaces) that provide relatively wide and direct access on system
on which it runs.\ \gls{cli} shells are \gls{cli} programs that allow users to
execute commands and other \gls{cli} utilities\cite{man_man,linfo_terminal}.

\textbf{\gls{cli} utilities} are programs that are executed from
within the \gls{cli} environments. Some popular examples of shells
and \gls{cli} utilities are mentioned in Table\ \ref{t:utils}.

\begin{table}
  \caption{Examples of shells and \gls{cli} utilities}\label{t:utils}
  \begin{tabular}{p{3cm}|p{4cm}|p{8cm}}
    Name & System & Description \\
    \hline
    Bash, Zsh
    & Linux, macOS
    & Popular Unix-like shells, compliant with POSIX standard.\\
    Cisco IOS Shell
    & Cisco devices
    & Shell used in Cisco network devices. \\
    bluetoothctl
    & Linux, macOS
    & \gls{cli} utility for managing Bluetooth devices.\\
    git
    & Cross-platform
    & \gls{cli} utility for version control using Git system. \\
    nmcli (NetworkManager)
    & Linux
    & Application for controlling network settings. \\
  \end{tabular}\\
\end{table}

\subsubsection{\gls{tui} Applications}

\glspl{tui} are a type of user interface that is a
further extension of \glspl{cli}. They provide a more visually structured and
intuitive way for users to interact with applications by using libraries such
as ncurses, which use the \textbf{terminal capabilities} emulators to create
rich text-based interfaces.\\

\textbf{Terminal capabilities} is a special functionality of the terminal, which
gives the application more control over it's output (i.e.\ move cursor in all
  directions, change the color of the output and rewrite/clean
characters)\cite{man_man,linfo_terminal}.\\

Some examples of applications using \gls{tui} are mentioned in
Table\ \ref{t:utils2}.
\begin{table}
  \caption{Examples of \gls{tui} utilities and programs}\label{t:utils2}
  \begin{tabular}{p{3cm}|p{5cm}|p{8cm}}
    Name & System & Description \\
    \hline
    vi/vim/nvim
    & Linux, macOS, Windows
    & \gls{tui} visual
    text editors. \\
    mc (Midnight commander)
    & Linux, macOS, Windows
    & \gls{tui} file explorer.\\
    htop
    & Linux
    & Task manager.\\
    nmtui (NetworkManager)
    & Linux
    & Application for controlling network settings. \\
  \end{tabular}\\
\end{table}

\subsection{Document Representation}

This section provides the background on the techniques for representing
documents in a way that allows for accurate retrieval. It also describes the
techniques for generating the semantic vector representations of documents,as
well as the storage and retrieval of these representations using vector
databases.

\subsubsection{Text Summarization}

\textbf{Text summarization} is defined as the process of a short, accurate, and
fluent summary of a longer text document. This technique is extensively used in
various applications, such as intent-classification, question-answering, and
information retrieval\cite{zhang2024systematicsurveytextsummarization}.

For the purpose of this thesis, we will focus on the single-document
summarization, with generation of semantic vector representation of
the document as the main goal.\\

\subsubsection{Semantic Document Search and Retrieval}\label{emb:semvec}

Currently, the semantic representation of
documents is generally categorized into two primary
groups\cite{khan2026retrieval}:

\begin{itemize}
  \item \textbf{Sparse vector representations (sparse retrieval):} This
    method mostly relies on traditional keyword-based approaches,
    such as \gls{bow}, one-hot encoding, \gls{lsindexing} or
    \gls{tf-idf} to retrieve documents based on the presence or
    quantity of specific terms. It is computationally efficient and
    works well for simple queries, but it may struggle with complex
    queries or those that require understanding of context and
    semantics.\\ \textsl{The common example of sparse retrieval is
      \gls{bow} model, which represents documents as vectors of word
      counts. It needs the dictionary of all possible words (or subset
      of them) to be defined. Then, the size of the vector is equal to
      the size of the dictionary, and each element position corresponds
    to the word in the dictionary.}

  \item \textbf{Dense vector representations (dense retrieval):} This method
    uses dense vector representations of documents and queries
    to calculate similarity. It typically involves training a neural
    network to generate embeddings that capture the semantic meaning of the
    text. The similarity between the query and documents is then calculated
    using cosine similarity or other distance metrics. Dense retrieval can
    be more effective for complex queries and can capture nuances in
    language, but it may require more computational resources compared to
    traditional sparse retrieval.
\end{itemize}

The original article also mentions \textbf{hybrid retrieval techniques},
that aim to combine the computational efficiency of sparse methods
with the ability of dense representations to alleviate the vocabulary
mismatch problem described in the article. These techniques typically
utilize \textbf{document expansion} to bridge the gap between exact
lexical matches and semantic meaning.
\textsl{For example, the
  SPLADE model expands the sparse representation of a document or query
  with semantically related terms, allowing for better retrieval
performance even when exact keywords are missing..}
The thesis is not going to focus on the hybrid retrieval techniques,
but it is important to acknowledge the existence of such techniques,
as they can be used in the future research to further improve the
retrieval performance of the system.\\

\subsubsection{Dense Representation Vector Generation}

\textbf{Dense representation vector generation} is also an important
aspect of the representation of documents. As the embedding
layer\ref{sec:emb-layer} of either dense or sparse retrieval
represents the document as a collection of vectors, one for each
token, it is important to have the techniques to generate the single
vector that represents this whole collection.

\begin{itemize}
  \item \textbf{Classical Approaches:} Theese techniques include
    methods like averaging, max pooling and other simple
    aggregation methods to combine the token embeddings into a
    single vector. These methods are computationally efficient
    and easy to implement, but they may not capture the full
    semantic meaning of the document, but rather features like
    frequency of words or presence and relationships based on it,
    but not the context.
  \item \textbf{Neural Network-based Approaches:} These techniques involve
    training a neural network, such as a transformer (embedding
    model, mentioned in\ \ref{sec:embedding-models}) or recurrent
    neural network, to generate a single vector representation for
    the entire document. The possible approaches are:
    \begin{itemize}
      \item \textbf{Multi-head attention-based summarization:} This method
        uses a transformer architecture retrieve weights for each token
        based on its position, relationships with other tokens and
        context, normalizes them, and then combines the token
        embeddings into a single vector using these weights. This
        approach can capture complex relationships between tokens and
        provide a more accurate representation of the document's
        meaning\cite{sharma2024unmacapsumtunifiedmultiheadattentiondriven}.
      \item \textbf{[CLS] token-based summarization:} This method uses a
        special token (e.g., [CLS] in \gls{bert}) that is added to the
        beginning of the document. The transformer is trained to update
        the values of this token based on the context of the entire
        document, and the final value of this token is used as the
        summary vector for the
        document\cite{behrendt2025maxpoolbertenhancingbertclassification,devlin2018bert}.
    \end{itemize}
    \textsl{It is important to note that to increase the accuracy of the
    summarization, it is often necessary to have single-topic documents}
\end{itemize}

\subsubsection{Vector Databases}\label{sec:vector-db}

\textbf{Vector databases} are specialized databases designed to store and
manage high-dimensional vector data and perform efficient similarity search
based on different metrics (e.g., cosine similarity, L2 or Euclidean distance).
Examples of vector databases include FAISS\cite{faiss2026},
Annoy\cite{annoy2026}, and Milvus\cite{milvus2026}.

For the purpose of this research, we will be using dense retrieval, as it is
more suitable for complex queries and can capture the semantic meaning of the
documents, which is more relevant for the task of choosing the right document
based on the user query.

\subsection{Large Language Models}

This section provides a brief overview of the aspects of \glspl{llm}, relevant
for this thesis. It describes the architecture, types, limitations of
\gls{llm}, the main protocols for the interaction, as well as the advanced
concepts.

\subsubsection{Language Models}

Language models are mathematical models designed to predict and generate
sequences of text. They are crucial in artificial intelligence for performing
various tasks like language translation, summarization, and content creation.
Over time, they have evolved significantly, powered by advances in neural
networks\cite{naveed2023}.

\subsubsection{\gls{llm} Overview}

\glspl{llm} refer to transformer language models that contain hundreds of
billions (or more) of parameters, which are trained on massive text data, such
as \gls{gpt}, \gls{bert}, \gls{lstm} or Mamba.\@ \glspl{llm} exhibit strong
capacities to understand natural language and solve complex tasks (via text
generation)\cite{radford2018improving,devlin2018bert,touvron2023llama}.

The architecture of \glspl{llm} can be divided into several main components:
\begin{itemize}
  \item \textbf{Tokenizer}: This component is responsible for providing the
    representation of the input text in a format that can be processed by
    the model. It breaks the input text into smaller units, to be able to
    represent it as vectors and then process it. The set of possible tokens
    is called the \textbf{vocabulary}, and can include words, subwords, or
    even individual characters. It is important to note that the vocabulary
    is fixed after training, which means that any new or rare words or
    combinations of characters (for example, command flags or arguments) may
    not be processed accurately or at all.
  \item \textbf{Embedding Layer}\label{sec:emb-layer}: This layer
    converts the tokenized input
    into dense vector representations, which capture the semantic meaning
    of each token. There are two main types of embeddings: \textbf{static}
    and \textbf{positional}. Static embeddings, such as Word2Vec or GloVe,
    assign a fixed vector to each token, regardless of the context in which
    it appears. Positional embeddings, on the other hand, generate
    different vectors for the same token based on its context in the
    input text, allowing for a more nuanced understanding of
    language\cite{mikolov2013distributed,pennington2014glove,zhao2023}.
  \item \textbf{Transformer Layers}: These layers are the core of the \gls{llm}
    architecture, responsible for processing the input and generating the
    output. They consist of multiple attention~\cite{attention} heads and
    feedforward networks that allow the model to capture complex
    relationships between tokens and generate coherent and contextually
    relevant responses.
  \item \textbf{Output Layer}: This layer generates the final output of the
    model, which can be in the form of text, probabilities, or other formats
    depending on the task. It typically consists of a linear layer followed by
    a softmax activation function to produce a probability distribution over the
    vocabulary for text generation tasks.
\end{itemize}

As mentioned above, \glspl{llm} are mainly built upon the transformer
architecture~\cite{attention}, where \textbf{multi-head attention layers} are
stacked in a very deep neural network. They are based on the transformer
architecture.\\

In the original article, previously mentioned transformers
consist of two main parts~--- \textbf{Encoder} and \textbf{Decoder}.
The usage of the architecture based on attention implies that the
model generates the
output text sequentially, one token at a time, while attending to all
previously generated tokens. This can be a limitation for long prompts, as the
model can only attend to a limited number of previous tokens (this limit is
called \textbf{context length} or \textbf{window size}). Furthermore, the
longer the prompt, the more computationally expensive it becomes to generate
the output, as the model has to process all previous tokens for each new
token\cite{attention}.\\

\textbf{Encoder}: The encoder maps an input sequence of symbols (tokens) to a
sequence of continuous representations. The encoder is composed of a stack of a
number of identical layers. Each layer has two sub-layers. The first is a
multi-head self-attention mechanism, and the second is a simple, position-wise
fully connected feed-forward network\cite{attention}.\\

\textbf{Decoder}: The decoder then generates an output sequence of symbols
(tokens) one element at a time. At each step the model is auto-regressive,
consuming the previously generated symbols as additional input when generating
the next. The decoder is also composed of a stack of a number of identical
layers. In addition to the two sub-layers in each encoder layer, the decoder
inserts a third sub-layer, which performs multi-head attention over the output
of the encoder stack\cite{attention}.\\

For the purpose of this thesis, it is enough to know that nowadays \glspl{llm}
are mainly based on Decoder-only architecture, which means they only use the
Decoder part of the original transformer architecture. Additionally,
Decoder-only models seem to be typically easier to train and fine-tune, as they
require less complex training data and objectives compared to Encoder-Decoder
models. That means that on the same numbers of training tokens, Decoder-only
models can achieve better performance than Encoder-Decoder
models\cite{zhao2023,wei2023chainofthoughtpromptingelicitsreasoning}.\\

\subsubsection{Embedding \gls{llm}}\label{sec:embedding-models}

\textbf{Embedding models} are a type of neural network architecture
that aims to
extract features from text with the natural languages (including words,
sentences, and documents)\cite{nie2025textembeddingmeetslarge}.

An \textbf{Embedding \gls{llm}} is a embedding model that is
specifically designed
to generate dense vector representations (embeddings) of content (either text,
audio, video or other types of data). These embeddings capture the semantic
meaning of the content, allowing for tasks such as semantic search, clustering,
and other applications that require understanding of the content's meaning.\\

The advantage of using embedding \gls{llm} in comparison to traditional
embedding models is that they can capture more complex relationships between
tokens and provide a more accurate representation of the content's meaning. This
is because embedding \gls{llm} are typically based on transformer architectures,
which use attention mechanisms to weigh the importance of different tokens in
the input text, allowing them to capture context and relationships between
tokens more effectively than traditional embedding
models\cite{zhang2024systematicsurveytextsummarization,khan2026retrieval}.\\

\subsubsection{\gls{moe} Extension}

While the architectures above define how information flows through the layers,
the \gls{moe} is an important extension that changes how the feedforward layers
operate, with the goal to increase scalability and efficiency.

In a traditional dense transformer, which can be reffered to as \textbf{Dense},
the whole decoder is a monolith, meaning that while processing, every neuron is
used. The \gls{moe} architecture splits the massive, singular
feedforward network into multiple smaller, quasi-specialized (as everything is
learned) \textsl{expert} subnetworks and introduces a \textsl{router} (or
gating network) to orchestrate them. The router acts as a traffic controller,
dynamically selecting only a tiny subset of experts~--- typically one
or two~---to
process it.

When it comes to comparing \gls{moe} and dense models in scope of
smaller devices
with certain limitations, \gls{moe} architectures introduce the
changes in two main
aspects compared to dense models: computational efficiency and memory usage.
\begin{itemize}
  \item \textbf{Computational Efficiency:} \gls{moe} models are more
    computationally
    lighter as the selective activation leads to significant reductions in
    computation time and energy consumption. However, such step makes the
    model less intelligent, especially on smaller scales. In contrary, a
    dense model involves every parameter in the computation for each token,
    leading to higher computational costs, but also more consistent performance.
  \item \textbf{Memory Usage:} The major drawback of \gls{moe} models
    is their high
    memory requirements. Since all experts must be loaded into memory to allow
    the router to select them dynamically, the total memory footprint is
    much larger than that of a dense model. Additionally, such architecture
    allows for knowledge duplication, thus less efficient memory usage.
\end{itemize}
Overall, \gls{moe} models tend to be more efficient in terms of
computation, but have
worse performance over dense models with same number of parameters. However, if
VRAM is not a constraint, MoE models can achieve better performance than dense
models within the same computational budget.

\subsubsection{OpenAI API}

The OpenAI protocol refers to the set of APIs and communication standards
developed by OpenAI for interacting with their AI models. It typically uses
RESTful HTTP endpoints with JSON payloads for requests and responses.
Authentication is handled via API keys, and communication is secured over
HTTPS. The protocol supports various endpoints for tasks such as completions,
chat, embeddings, and file uploads, enabling integration with a wide range of
applications. More specific information can be found in OpenAI API
Documentation\ \cite{openai}.

\subsection{AI Agent Systems}\label{agent}

The evolution of \glspl{llm} has progressed beyond simple text generation to the
concept of \textbf{AI Agents}, which utilize \glspl{llm} with the
capacity to plan,
act, and interact with external environments. AI agents are autonomous
entities, capable of planning, decision-making, and performing actions to
achieve complex goals\cite{naveed2023,zhao2023}.\\

Two major changes that AI Agents bring compared to traditional \glspl{llm} are:
\begin{itemize}
  \item \textbf{Tool Use:} AI Agents can leverage external tools,
    RAG pipelines or APIs to retrieve information or perform actions
    beyond text generation.
  \item \textbf{Iterative Generation:} AI
    Agents can generate multiple steps of reasoning or actions iteratively,
    allowing them to break down complex tasks into manageable sub-tasks.
\end{itemize}

The secound point is especially important for the purpose of this
research, as it allows the agent to decompose complex tasks into
steps small enough to be handled by local \glspl{llm} with limited
capabilities. We can define the workflow of \gls{llm} Agents
~\cite{nvidia-llm-agent} as pseudocode in Listing\ \ref{lst:agent}.

\begin{lstlisting}[,
  caption={Pseudocode for \gls{llm} Agent workflow},
  label={lst:agent},
  language=pseudo,
  style=code-listing,
  float
]

function LLM_Agent(task, context):
    subtasks = LLM.decompose_task(task)  # Break down the main task into smaller sub-tasks
    for subtask in subtasks:
        switch subtask.type:
            case "tool":
                result = use_tool(subtask.tool, subtask.tool_input)
            case "further_decomposition":
                # Recursive call for further decomposition
                result = AI_Agent(subtask.task, context)
            case "generate":
                result = LLM.generate_response(subtask.task)
        results.append(result)
    final_output = LLM.summarize_results(results)
    return final_output

\end{lstlisting}

In essence, AI Agents perform tasks, dividing complex objectives into
manageable steps, and iteratively prompting themselves to execute
these steps using various tools.

\subsection{Retrieval-Augmented Generation (RAG)}

\textbf{\gls{rag}} is a technique that allows \glspl{llm}
to utilize external knowledge bases, significantly reducing hallucinations and
increasing accuracy\cite{zhao2023}.

We can define the workflow of \gls{rag} as follows as
pseudocode in Listing\ \ref{lst:rag}.

\begin{lstlisting}[,
  caption={Pseudocode for \gls{rag} workflow},
  label={lst:rag},
  language=pseudo,
  style=code-listing,
  float
]

function LLM_RAG_generate(user_query, LLM, document_database):
    # Step 1: Retrieve relevant documents from the document database
    relevant_documents = retrieve_documents(document_database, user_query)

    # Step 2: Inject retrieved documents into the input prompt for the LLM
    augmented_prompt = construct_prompt(user_query, relevant_documents)

    # Step 3: Pass the augmented prompt to the LLM to generate a response
    llm_response = LLM.generate_response(augmented_prompt)

    # Step 4: Return the final response
    return llm_response

\end{lstlisting}

This technique involves document retrieval, the best general approach for which
is to use
\textbf{semantic document search and retrieval}\ref{emb:semvec}. By this, we
mean the set of techniques of indexing and searching documents based on their
semantic representations. Therefore, there are two main components for the
retrieval system: document representation generator (embedding model, mentioned
in Section\ \ref{sec:embedding-models}) and storage with searching capabilities
suited for the task (see \ref{sec:vector-db}).\\

\paragraph{Different Granularities In Retrieval}

\gls{rag} systems can retrieve information at multiple granularities, for
example:

\begin{itemize}
  \item \textbf{Document-level retrieval}
  \item \textbf{Passage-level retrieval}
  \item \textbf{Sentence-level retrieval}
\end{itemize}

The choice of granularity defines the smallest unit, for which the (vector)
representation is generated. This affects the computational load for the
\gls{rag} system for the creation and retrieval, or how much different units can
be fit into \gls{llm} context window or answer size.

\paragraph{Query Types}

%  Level-1
%
% Explicit Facts: These queries are asking about explicit facts
% directly present in the given data without requiring any additional
% reasoning. This is the simplest form of query, where the model’s
% task is primarily to locate and extract the relevant information.
% For example, "Where will the 2024 Summer Olympics be held?" targets
% a fact contained in the external data.
% Level-2
%
% Implicit Facts: These queries ask about implicit facts in the data,
% which are not immediately obvious and may require some level of
% common sense reasoning or basic logical deductions. The necessary
% information might be spread across multiple segments or require
% simple inferencing. For instance, the question "What is the
% majority party now in the country where Canberra is located?" can
% be answered by combining the fact that Canberra is in Australia
% with the information about the current majority party in Australia.
% Level-3
%
% Interpretable Rationales: These queries demand not only a grasp of
% the factual content but also the capacity to comprehend and apply
% domain-specific rationales that are integral to the data’s context.
% These rationales are often explicitly provided in external
% resources and is typically not present or rarely encountered during
% the pre-training phase of a general large language model. For
% example, in the realm of pharmaceuticals, an LLM must interpret FDA
% Guidance1 documents—which represent the FDA’s current thinking—to
% evaluate whether a specific drug application adheres to regulatory
% requirements. Similarly, in customer support scenarios, the LLM
% must navigate the intricacies of a predefined workflow to process
% user inquiries effectively. In the medical field, many diagnostic
% manuals provide authoritative and standardized diagnostic criteria,
% such as management guidelines for patients with acute chest pain
% [14]. By effectively following these external rationales, it is
% possible to develop a specialized LLM expert system for managing
% chest pain. This involves understanding the procedural steps and
% decision trees that guide a support agent’s interactions with
% customers, ensuring responses are not only accurate but also comply
% with the company’s service standards and protocols.
% Level-4
%
% Hidden Rationales: This category of queries delves into the more
% challenging realm where the rationales are not explicitly
% documented but must be inferred from patterns and outcomes observed
% in external data. The hidden rationales here refer not only to the
% implicit reasoning chains and logical relationships, but also to
% the inherently challenging and non-trivial task of identifying and
% extracting the external rationales required for each specific
% query. In IT operational scenarios, for example, a cloud operations
% team may have addressed numerous incidents in the past, each with
% its own unique set of circumstances and resolutions. The LLM must
% be adept at mining this rich repository of tacit knowledge to
% discern the implicit strategies and decision-making processes that
% were successful. Similarly, in software development, the debugging
% history of previous bugs can provide a wealth of implicit insights.
% While the step-by-step rationale for each debugging decision may
% not be systematically recorded, the LLM must be capable of
% extracting the underlying principles that guided those decisions.
% By synthesizing these hidden rationales, the LLM can generate
% responses that are not only accurate but also reflective of the
% unspoken expertise and problem-solving approaches that have been
% honed over time by experienced professionals.

We can generally classify queries that user can ask to the \gls{rag} system into
four main types, defined in the original article\
\cite{zhao2024retrievalaugmentedgenerationrag}:

\begin{itemize}
  \item \textbf{Level-1: Explicit Facts:} These queries are asking
    about explicit facts
    directly present in the given data without requiring any additional
    reasoning. This is the simplest form of query, where the model’s task is
    primarily to locate and extract the relevant information.
  \item \textbf{Level-2: Implicit Facts:} These queries ask about implicit
    facts in the data, usually searching for the information in multiple
    segments and require some level of common sense reasoning or basic logical
    deductions.
  \item \textbf{Level-3: Interpretable Rationales:} To answer these queries,
    the model needs to not only retrieve and understand the facts in the
    database, but also suit the answer to the specific domain, by
    understanding and applying the domain-specific rationales.
  \item \textbf{Level-4: Hidden Rationales:} The answer to these queries is
    usually not explicitly documented, but must be deduced from patterns and
    outcomes found in external data.
\end{itemize}

With such definition of query types, it is possible to map the responsibilities
of the parts of the \gls{rag} system to the query types.\\

Level-1 queries can be mostly handled by the retrieval component, with the role
of the \gls{llm} being to filter the relevant information from the retrieved
documents.\\
Level-2 queries require the \gls{llm} to perform some reasoning and inference,
to deduct Level-1 queries from the Level-2 to ask the retrieval component, and
then combine the retrieved information to produce the final answer.\\
Level-3 queries require the \gls{llm} to understand the domain-specific
rationales, usually to be pre-trained on the documents containing these
rationales, and then apply them to the retrieved information to produce the
final answer. Along with Level-4 it requires advanced reasoning capabilities.\\

In this thesis, we will be focusing on the Level-1 and Level-2 queries, as they
are more relevant for the task of choosing the right document based on the user
query, and the application of the system in the context of operating on
manuals usually requires the retrieval of explicit and implicit facts, rather
than the deduction of hidden rationales.

\subsubsection{Common RAG Testing Techniques}

This subsecion describes most common RAG Testing Techniques, essential for the
evaluation of the performance of \gls{rag} systems\cite{rag-evaluation}.

\paragraph{Precision@k}

\textbf{Precision@k} measures the proportion of relevant documents or passages
that appear within the top-k retrieved results. It is defined as:

\[
  \text{Precision@k} = \frac{\text{number of relevant documents in
  top-k}}{\text{k}}
\]

This metric directly assesses the quality of the retrieval component. For
example, Precision@1 measures whether the most relevant document is ranked
first, while Precision@5 measures whether relevant documents appear within the
top 5 results. Higher precision at low k values (e.g., k=1) is generally
preferred, as it indicates the system's top results are highly relevant.

\paragraph{Recall@k}

\textbf{Recall@k} measures the proportion of all relevant documents that are
successfully retrieved within the top-k results:

\[
  \text{Recall@k} = \frac{\text{number of relevant documents
  retrieved in top-k}}{\text{total number of relevant documents}}
\]

Recall provides a complementary perspective to precision: a system can achieve
high precision by returning few results, but recall indicates whether the system
finds a significant portion of all relevant documents. Recall@k is particularly
important when comprehensive coverage is needed.

\paragraph{Mean Reciprocal Rank (MRR)}

\textbf{Mean Reciprocal Rank} measures the average rank of the first relevant
result across all queries:

\[
  \text{MRR} = \frac{1}{|Q|} \sum_{i=1}^{|Q|} \frac{1}{\text{rank}_i}
\]

where \textsl{rank\textsubscript{i}} is the position of the first relevant
document for query i, and |Q| is the total number of queries. MRR rewards
systems that find relevant documents early in the ranking and is particularly
useful when users typically only examine the top few results.

\paragraph{Normalized Discounted Cumulative Gain (NDCG)}

\textbf{Normalized Discounted Cumulative Gain} combines relevance judgments with
ranking position to account for the fact that lower-ranked relevant documents
are less valuable than higher-ranked ones:

\[
  \text{DCG} = \sum_{i=1}^{k} \frac{\text{rel}_i}{\log_2(i+1)}
\]

where \textsl{rel\textsubscript{i}} is the relevance score of the document at
position i. \gls{ndcg} normalizes this by the ideal ranking (IDCG) to produce a
score between 0 and 1. NDCG is valuable for assessing ranking quality when there
are multiple levels of relevance (e.g., highly relevant, partially relevant, not
relevant).

\paragraph{F1 Score for Retrieval}

For binary relevance judgments (relevant or not relevant), the \textbf{F1 score}
combines precision and recall:

\[
  \text{F1} = 2 \times \frac{\text{Precision} \times
  \text{Recall}}{\text{Precision} + \text{Recall}}
\]

The F1 score is useful for balancing the trade-off between precision and recall
when a single metric is needed to optimize system performance.

\subsection{Linux Manual Pages}

In this thesis and implementation, we focus on processing linux man manual pages
as data for testing and validation and by the term \textbf{document} we will
mean a man manual page.

In this part we provide the description of the main conventions for
man pages, which are relevant for the solution.

man is the system's manual pager. Each page argument given to man is normally
the name of a program, utility or function. The manual page associated with
each of these arguments is then found and displayed. A section, if provided,
will direct man to look only in that section of the manual\cite{man_man}.

Conventionally, manual pages are divided into several groups, which are listed
in\ Table \ref{t:man-groups}.
A manual page consists of several sections. Conventional section
names are listed in\ \ref{t:man-sections}. The most important sections for the
purpose of this thesis are \textbf{DESCRIPTION}, which provides a detailed
description of the program, its functionality and usage, and \textbf{OPTIONS},
which lists the flags and arguments that can be used with the program, along
with their descriptions.

\begin{table}
  \caption{Conventional man page groups}\label{t:man-groups}
  \small
  \begin{multicols}{2}
    \begin{enumerate}
      \item Executable programs or shell commands
      \item System calls (functions provided by the kernel)
      \item Library calls (functions within program libraries)
      \item Special files (usually found in /dev)
      \item File formats and conventions, e.g. /etc/passwd
      \item Games
      \item Miscellaneous (including macro packages and conventions),
        e.g. man(7), groff(7), man-pages(7)
      \item System administration commands (usually only for root)
      \item Kernel routines [Non standard]
    \end{enumerate}
  \end{multicols}
\end{table}

\begin{table}
  \caption{Conventional section names in man pages}\label{t:man-sections}
  \small
  \begin{multicols}{2}
    \begin{itemize}
      \item NAME
      \item SYNOPSIS (usage)
      \item CONFIGURATION
      \item DESCRIPTION
      \item OPTIONS (flags)
      \item EXIT STATUS
      \item RETURN VALUE
      \item ERRORS
      \item ENVIRONMENT
      \item FILES
      \item VERSIONS
      \item STANDARDS
      \item NOTES
      \item BUGS
      \item EXAMPLE
      \item AUTHORS
      \item SEE ALSO
    \end{itemize}
  \end{multicols}
\end{table}
