
\subsection{Existing Solutions Overview}

This section serves as listing of existing solutions relevant to the
evolution and formation of the thesis.
In Section\ \ref{sec:design-evolution} we will be reffering to the
solutions described
here, as well as to the key insights gained from their analysis.

\paragraph{Butterfish}

Butterfish is a shell companion that integrates \gls{llm}
capabilities directly into
interactive shell environments. It provides features such as command
suggestions,
error explanation, and context-aware assistance by streaming \gls{llm} responses
into the terminal. It relies on remote \glspl{llm} via the OpenAI API
and does not
support local \glspl{llm} execution.

\paragraph{llm-zsh-plugin}

The llm-zsh-plugin is a Zsh plugin that extends shell functionality
with \gls{llm}
integration for command generation and suggestions. It provides shell completion
and command assistance features but is dependent on remote \glspl{llm} services
and lacks support for offline, local model execution.

\paragraph{Zsh Shell Integration (Kruger approach)}

Another solution was proposed in article by Kruger\cite{kruger-web}, which
describes a setup for zsh shell with \gls{llm} features, such as command
suggestions, autocompletion, and in-terminal assistance. It uses OpenAI API to
access remote \glspl{llm}. This approach demonstrates how shell hooks can be
leveraged for \gls{llm} integration, but shares the limitation of requiring
external API access.

\paragraph{Warp Terminal}

This is a developement-oriented and general-purpose \gls{ai}-enabled terminal
emulator, which uses remote Agent \gls{llm} System to enchance the experience of
users. The most of its code remains closed-source. It is not adapted for
executing on small \glspl{llm} and it provides no support for local \glspl{llm}.
Additionally, it is not a \gls{cli}/\gls{tui} application, but a full
graphical terminal emulator.

This solution is popular and comercially successful, but it does not
completely fit into the scope of our work, as it is not designed to work with
local \glspl{llm}. However, it is worth mentioning, as it provides good AI-based
services for terminal users\cite{warp-web,warp-docs}.

\paragraph{gptme}

This is general-purpose \gls{cli} utility with bread \gls{ai}
functionality. It is compatible with OpenAI \gls{api}, which allows
it to use both remote and local \glspl{llm}. It uses Agentic
architecture and implements \gls{mcp} protocol. This enables it to
use various tools (terminal, browser, code execution, etc.). It is
also context-aware, which means it can read and understand the files
in the current working directory and acknowledge host
\gls{os}\cite{gptme-web,gptme-docs}. Example of the interaction with
gptme is provided in Listing\ \ref{lst:gptme}.

\begin{lstlisting}[,
  caption={Example of gptme interaction with browser tool},
  label={lst:gptme},
  language=bash,
  style=code-listing,
  float
]
\$ gptme --tools=browser "How many calories are there in one spoon of sugar?"
[11:07:21] WARNING Skipping all confirmation prompts. cli.py:186
[11:07:21] Using browser tool with lynx
Found openai API key, using openai provider
Using model: openai/gpt-5
[11:07:22] Using logdir: ~/.local/share/gptme/logs/2025-12-10-running-sleepy-robot
Using workspace: ~/Workspace/bachelor/project/document
Skipped 2 hidden system messages, show with --show-hidden
--- ^^^ past messages ^^^ ---
User: How many calories are there in one spoon of sugar?
Assistant: Short answer: about 16 calories per level teaspoon.

Details:
- Level teaspoon (≈4 g sugar): ~15–16 kcal
- Heaped/rounded teaspoon (≈5–7 g): ~20–28 kcal
- Tablespoon (≈12–15 g): ~48–60 kcal

Sugar has ~4 kcal per gram (sucrose ~3.87 kcal/g), so you can estimate by weight if needed.
\end{lstlisting}

Key capabilities:
\begin{itemize}
  \item Uses OpenAI API for accessing remote/local \glspl{llm}
  \item Uses Agent architecture and MCP tools
  \item Is context-aware and is able to use tools (terminal, browser,
    code execution, etc.)
  \item Implements MCP protocol
\end{itemize}

When tested manually with local \glspl{llm} (Gemma3 4b
\ \cite{gemma-web}, Qwen2.5 0.5B/1.5B\ \cite{qwen-web}), it performed quite
poorly, as the models were not able to handle the complexity of the
tasks. It usually due to too long prompts which made \glspl{llm}
halucinate on too many details. This shows, that the solution is not
fully adapted for local \glspl{llm}, which is one of the main goals
of our work. This should be considered when designing our solution to
use more fine-grained prompts for local \glspl{llm} and optimize the
architecture for local \glspl{llm}.

\paragraph{gemini-cli}

This a is general-purpose \gls{cli} chat/agent \gls{tui} application
by Google. It uses Gemini \glspl{llm} to provide \gls{ai}
capabilities in terminal. It is useful for various tasks, including
code generation, data analysis, and general Q\&A. It also supports
\gls{mcp} protocol and custom tool integration. Hovewer, it is
dependable on remote \glspl{llm} and cannot be used with local
\glspl{llm}\cite{gemini-web,gemini-docs}.

\paragraph{Yai (Your \gls{ai} Assistent)}

This application is an interactive \gls{tui} chat application, which
uses OpenAI to
provide user with \gls{ai} capabilities in terminal. It supports
tools interaction
and basic agentic architecture. In current form, it is not able to use local
\glspl{llm}. Hovewer, it is possible to adapt it for local \gls{llm}s
with some modifications
in its source code.\cite{yai-web,yai-docs}.

\paragraph{Aider}

This is an \gls{llm}-powered code agent with focus on software
development tasks and
terminal integration. It supports various tools and is context- and
project-aware. It is able to use both remote and local \glspl{llm}. It is also
open-source.\cite{aider-web,aider-docs}.

\paragraph{termai}

This is a terminal-based \gls{llm} assistant similar to gptme,
written in Rust. It
supports various \gls{llm} providers, but it is not able to use local
\glspl{llm} in its
current form. It is open-source and has basic agentic
architecture\cite{termai-web}.

\paragraph{AiTerm}

The application that utilizes tmux terminal multiplexer to access the
terminal \gls{io}. It sends the terminal output as context to chosen
\gls{llm} and gets the response back. It has execution agent
capabilities, which allows it to run commands in terminal based on
\gls{llm} response. It is open-source and is adapted to use with
OpenAI \gls{api}, thus and it can be used with local \glspl{llm} as well.

We can draw inspiration from its approach to Bash integration using
tmux sessions to provide seamless interaction between \gls{llm} and
terminal. Hovewer, this approach turned out to be somewhat clunky in
practice, as it couldn't handle well the ~\gls{tui} applications, but
only get bash's responses\cite{aiterm-web}.

\paragraph{Other solutions}

There are also such solutions, like \textsl{ai-terminal-assistent (github)}{}
or \textsl{cmd-ai (github)}, which are just wrappers for Rest \gls{api} calls to
remote \glspl{llm}. They provide basic \gls{ai} capabilities in terminal, but
have no significant features.

\subsection{Technology \& Design Evolution}\label{sec:design-evolution}

The scope of this thesis has changed through multiple iterations as existing
solutions were found and analyzed and either already existing solutions were
found, or practical constraints become apparent. This section documents the
technical decisions and trade-offs that led to the final architecture. We will
go through the key iterations in the design evolution, highlighting the
reasoning behind each shift and key knowledge gained from testing and analysis
of existing solutions.

\subsubsection{Iteration 1: External LLM with Broad Features}
The initial vision for the project was ambitious: an AI-powered shell
wrapper with
broad capabilities including command autocompletion, suggestions, in-terminal
assistance, and dangerous command detection/prevention. However, numerous
existing solutions % mention via references relevant tools
already provided similar features, relying on remote LLMs,
which were already tested in this space.
\paragraph{Limitation:} existing numerous solutions.
\paragraph{Gained Knowledge:} overall overview of the field and
existing solutions,
OpenAI API.

\subsubsection{Iteration 2: Local-First Architecture with Modular Design}
Next iteration of the software was seen as constraint to local \glspl{llm} only,
while trying to maintain the same broad feature set. However, early testing with
small local \glspl{llm} revealed significant limitations in knowledge breadth,
high hallucination rates, and showed that \glspl{llm} struggled with
classification and reasoning tasks required for features like dangerous command
detection.

We had to significantly narrow the scope of features to those that could be
reliably implemented with local \glspl{llm} and empirically tested. This meant
dropping features like:
\begin{itemize}
  \item \textbf{Seamless integration with \texttt{bash} or other shells} ---
    While shell integration through tools like tmux (as demonstrated by AiTerm)
    and zsh plugins appeared promising, this approach proved
    difficult to implement
    reliably. Testing showed three main challenges: (1) basic command execution
    and user interaction were hard to verify systematically, (2) distinguishing
    between natural language queries and bash commands showed low
    persistence and
    reliability without a dedicated classification module, and (3) handling I/O
    streams and terminal state significantly increased system complexity. These
    challenges suggest shell integration requires more sophisticated
    architecture
    than initially apparent.

  \item \textbf{Dangerous command detection} --- While conceptually appealing,
    empirical validation of dangerous command detection proved infeasible within
    the thesis scope. The primary challenge was creating a
    comprehensive verification
    system and dataset to test the classifier reliably. Additionally,
    defining what
    constitutes a "dangerous" command depends heavily on user context and system
    configuration, making formal evaluation difficult. This feature
    was deprioritized
    in favor of more objectively measurable functionality.

  \item \textbf{Agentic systems with autonomous execution} --- AI agents capable
    of planning and executing complex tasks seemed attractive but
    posed significant
    testing challenges. Verifying agent behavior requires extensive
    integration testing
    and careful evaluation of edge cases. Within the thesis timeline and without
    a production deployment scenario, creating reliable tests for
    autonomous execution
    was not practical. This direction was deferred to future
    research, with potential
    connection to reinforcement learning and multi-step reasoning techniques.
\end{itemize}

\paragraph{Limitation:} local \glspl{llm} have limited capabilities for complex
reasoning and long context handling.
\paragraph{Gained Knowledge}: Through extensive testing with small local models
(Gemma 3 4B and Qwen2.5 0.5B/1.5B), we identified key constraints that shaped the
architecture. Local \glspl{llm} exhibited poor performance on:
\begin{itemize}
  \item Long prompts and multi-turn conversations (hallucination
      increases significantly
    with context length)
  \item Abstract reasoning and classification tasks (models lack
      sufficient training
    on narrow tasks like security classification)
  \item Complex chained reasoning requiring multiple logic steps
\end{itemize}
This led to discovering techniques for enhancing local \gls{llm}
performance within
these constraints: Retrieval-Augmented Generation (RAG) to reduce
reliance on model
knowledge, embedding models for efficient semantic search, and task
decomposition
to avoid complex reasoning chains.

\subsubsection{Iteration 3: Focused RAG System}

Final design: concentrating on RAG (Retrieval-Augmented Generation) using
embeddings from local \glspl{llm}. Classification and agent modules were found
to be too demanding for local models to execute reliably. RAG provides a clear,
empirically testable approach: semantic search over structured knowledge sources
(Linux man pages) coupled with \gls{llm}-assisted response generation. This
architecture plays to the strengths of local models---embedding generation and
content summarization---while avoiding their weaknesses in complex reasoning.
The system decomposes the assistant task into two simpler operations: (1)
retrieval of relevant documentation using semantic similarity, and (2)
generation of a user-friendly response by re-ranking and summarizing retrieved
content. This decomposition approach is further detailed in the subsequent
chapters, where the system architecture and implementation are described in
detail.

\subsubsection{Final Design: Information Retrieval with LLM Assistance}

\textbf{Local LLM Limitations} --- Early experiments with small
\glspl{llm} revealed that local models struggle with:
\begin{itemize}
  \item Complex multi-step reasoning tasks
  \item Long context windows (hallucination increases with prompt length)
  \item Abstract feature understanding (e.g., dangerous command classification)
\end{itemize}

\textbf{Architecture Design} --- The final RAG approach leverages strengths
of local \glspl{llm}:
\begin{itemize}
  \item \textbf{Embedding Generation} --- Local models excel at creating
    semantic embeddings, enabling offline vector similarity search
  \item \textbf{Response Generation} --- Local models can re-rank and
    summarize retrieved content, avoiding long reasoning chains
  \item \textbf{Offline Operation} --- No dependency on external APIs,
    enabling deployment on resource-constrained systems
\end{itemize}

\subsection{Summary Existing Solutions}\label{sec:existing-solutions-summary}

The following criteria are essential for evaluating existing solutions and
guiding our architectural decisions. These criteria directly influenced which
features were considered feasible to implement and what trade-offs were
necessary given the constraints of local \gls{llm} execution:

\begin{itemize}
  \item \textbf{Local \gls{llm} Support} --- Core requirement for our
    project goal: a minimalistic shell assistant that can run on
    resource-constrained environments without reliance on external
    cloud APIs or internet connectivity.

  \item \textbf{Type} --- Implementation approach: terminal emulator
    (replacement shell), \gls{cli} utility (command-line tool), or
    \gls{tui} application (text-based interface). Each approach has
    different implications for user experience and system integration.

  \item \textbf{Open Source} --- Essential for transparency,
    reproducibility, and extensibility in academic thesis work.
    Allows detailed analysis of implementation approaches and
    potential integration with our solution.

  \item \textbf{Context Awareness} --- Ability to understand the
    current working directory, access local files, and acknowledge
    host \gls{os} characteristics. This enables more intelligent and
    contextual assistance tailored to user's environment.

  \item \textbf{Agentic Architecture} --- Support for autonomous task
    execution, planning, and tool usage. Agentic systems can break
    down complex tasks into subtasks and execute them iteratively,
    improving reliability and user experience.

  \item \textbf{Tool Integration} --- Extensibility through external
    tools and protocols. Support for standards like \gls{mcp} (Model
    Context Protocol), bash integration, code execution, and other
    interfaces affects the solution's versatility.

  \item \textbf{Language} --- Implementation language relevant to
    maintenance, extension, and integration possibilities. Different
    languages have different performance characteristics and ecosystem support.
\end{itemize}

Table\ \ref{tab:solutions-summary} compares existing solutions across
the evaluation criteria
described above. Note that this comparison informed architectural decisions in
designing our solution, particularly regarding which capabilities are feasible
to implement with local \glspl{llm}.

\begin{table}
  \centering
  \caption{Summary of existing solutions across key evaluation criteria.}
  \small
  \begin{tabularx}{\textwidth}{lXXX}
    \toprule
    \textbf{Solution} & \textbf{Warp Terminal} & \textbf{gptme} &
    \textbf{gemini-cli} \\
    \midrule
    \textbf{Open Source} & No & Yes & No \\
    \textbf{Local \gls{llm} Support} & No & Yes (Limited) & No \\
    \textbf{Context Awareness} & Partial (Display) & Yes & Yes \\
    \textbf{Agentic Architecture} & Yes & Yes & Yes \\
    \textbf{Type} & Terminal Emulator & \gls{cli} Utility & \gls{cli} Chat \\
    \textbf{Tool Integration} & Limited & \gls{mcp} Protocol &
    \gls{mcp} Protocol \\
    \textbf{Language} & Proprietary & Python & Go \\
    \bottomrule
  \end{tabularx}

  \medskip

  \begin{tabularx}{\textwidth}{lXXXX}
    \toprule
    \textbf{Solution} & \textbf{Aider} & \textbf{Yai} &
    \textbf{termai} & \textbf{AiTerm} \\
    \midrule
    \textbf{Open Source} & Yes & Yes & Yes & Yes \\
    \textbf{Local \gls{llm} Support} & Yes & Limited & No & Yes \\
    \textbf{Context Awareness} & Yes & No & No & Partial \\
    \textbf{Agentic Architecture} & Yes & Partial & Yes & Yes \\
    \textbf{Type} & Code Agent & \gls{tui} Chat & \gls{cli} Chat &
    Terminal Wrapper \\
    \textbf{Tool Integration} & Code Execution & Limited & Basic &
    Bash Integration \\
    \textbf{Language} & Python & Go & Rust & Python \\
    \bottomrule
  \end{tabularx}

  \medskip

  \begin{tabularx}{\textwidth}{lXXX}
    \toprule
    \textbf{Solution} & \textbf{Butterfish} & \textbf{llm-zsh-plugin} &
    \textbf{Zsh Integration (Kruger)} \\
    \midrule
    \textbf{Open Source} & Yes & Yes & Yes \\
    \textbf{Local \gls{llm} Support} & No & No & No \\
    \textbf{Context Awareness} & Limited & Limited & Limited \\
    \textbf{Agentic Architecture} & No & No & No \\
    \textbf{Type} & Shell Companion & Zsh Plugin & Shell Integration \\
    \textbf{Tool Integration} & Limited & Limited & Shell Hooks \\
    \textbf{Language} & Python & Python & Zsh/Shell \\
    \bottomrule
  \end{tabularx}

\end{table}\label{tab:solutions-summary}

\subsection{Technology Stack}

This section provides an overview of the key technologies and tools used in
the implementation of the solution.

\subsubsection{Python}

Python was chosen as the primary implementation language for this thesis due to
its simplicity, readability, and extensive ecosystem of libraries for machine
learning, and devise integration. Among such libraries, we use \texttt{os},
\texttt{subprocess}, and \texttt{pathlib}.
% TODO: add more libraries, such as for testing, logging, etc.

\subsubsection{Ollama}

Ollama\cite{ollama-web} is a local LLM service that enables running large
language models on consumer hardware without requiring cloud API calls. For this
thesis, Ollama was chosen as the primary LLM runtime over in-code alternatives
for several reasons:

\begin{itemize}
  \item \textbf{Simplicity and Ease of Use} --- Ollama provides a
    straightforward command-line interface for downloading,
    installing, and managing LLMs. This eliminates the need to
    implement model loading and inference logic directly in the application.

  \item \textbf{Model Flexibility} --- Ollama supports a wide variety
    of open-source models with different sizes and capabilities,
    allowing for easy experimentation with different architectures
    and parameter counts.

  \item \textbf{OpenAI API Compatibility} --- Ollama provides an
    OpenAI-compatible API endpoint, making it possible to test with
    both local (via Ollama) and remote (via OpenAI) models without
    changing application code. This flexibility was valuable for
    development and testing.

  \item \textbf{Resource Efficiency} --- Ollama is designed for local
    execution and provides optimized inference, reducing memory
    footprint and latency compared to running models directly in
    Python or other application languages.
\end{itemize}

\subsubsection{\gls{annoy}}

For the RAG retrieval component, \gls{annoy}\cite{annoy2026} was selected
as the vector database for storing and retrieving document
embeddings. This choice was motivated by:

\begin{itemize}
  \item \textbf{Simplicity and efficiency} --- a minimal python API
    with no complex server setup or external dependencies, making it
    ideal for prototype development.

  \item \textbf{File-Oriented Storage} --- Annoy saves indices
    to disk files, allowing persistent storage of embeddings without
    requiring a running database server.

\end{itemize}

\subsubsection{\glspl{llm} and Embedding Models}

The thesis testing and development primarily utilized several small, publicly
available \glspl{llm} and embedding models, summarized in
Tables\ \ref{tab:llm-summary} and\ \ref{tab:embedding-summary}.

\begin{table}
  \centering
  \scriptsize
  \caption{Summary of \glspl{llm} used in experiments (parameters,
  context, and licensing).}
  \label{tab:llm-summary}
  \begin{tabularx}{\textwidth}{XlXXX}
    \toprule
    \textbf{Model} & \textbf{Params} & \textbf{Context} &
    \textbf{License} & \textbf{Training data (public disclosure)} \\
    \midrule
    Qwen2.5-0.5B-Instruct & 0.5B & 32K (up to 128K; 8K gen) & Apache 2.0 &
    Up to 18T tokens (family-level disclosure). \\
    Qwen2.5-1.5B-Instruct & 1.5B & 32K (up to 128K; 8K gen) & Apache 2.0 &
    Up to 18T tokens (family-level disclosure). \\
    Qwen3-0.6B & 0.6B & 32K & Apache 2.0 &
    \textasciitilde{}36T tokens; web + doc-like corpora + synthetic
    math/code (family-level disclosure). \\
    Qwen3-1.7B & 1.7B & 32K & Apache 2.0 &
    \textasciitilde{}36T tokens; web + doc-like corpora + synthetic
    math/code (family-level disclosure). \\
    Granite-3.1-1B-A400M-Instruct (\gls{moe}) & 1.3B (400M active) & 128K &
    Apache 2.0 & 10T tokens (\gls{moe}); SFT mix as above (high-level disclosure). \\
    Gemma 3 (Google) & 4B & 32K & Gemma Terms of Use &
    Token counts disclosed; broad web/code/math/image sources
    (high-level disclosure). \\
    \bottomrule
  \end{tabularx}
\end{table}

\begin{table}
  \centering
  \scriptsize
  \caption{Summary of embedding models used in experiments
  (dimension, context, and licensing).}
  \label{tab:embedding-summary}
  \begin{tabularx}{\textwidth}{XlllX}
    \toprule
    \textbf{Model} & \textbf{Params} & \textbf{Context} &
    \textbf{Dim} & \textbf{Notes} \\
    \midrule
    Qwen3-Embedding-0.6B & 0.6B & 32K & up to 1024 & Apache 2.0;
    supports user-defined dimensions (MRL). \\
    EmbeddingGemma-300M (Google) & 300M & 2048 & 768 & Gemma Terms of
    Use; can be truncated to smaller dims via MRL. \\
    granite-embedding-30m-english & 30M & 512 & 384 & Apache 2.0;
    English-only embedding model. \\
    granite-embedding-278m-multilingual & 278M & 512 & 768 & Apache
    2.0; multilingual embedding model (12 languages listed by IBM). \\
    \bottomrule
  \end{tabularx}
\end{table}
