
\subsection{Component 1: Natural Language vs. Shell Command Classifier}

\paragraph{Evaluation Approach}{}
The testing for this component focuses on evaluating the accuracy of input
classification across different local \gls{llm} models. A fixed, balanced dataset
is classified into natural language vs.\ shell command inputs (see\
\ref{sec:classifier-approaches} for details on the approaches).

\paragraph{Testing Collections}{}
For testing the classifier, the manually curated collection
\texttt{classifier-100} contains 100 inputs with a balanced mix of natural
language queries and shell command inputs. It is split as follows:
\begin{itemize}
  \item 50 natural language queries (e.g., ``How do I list files in a
    directory? '')
  \item 50 shell command inputs (e.g., ``ls -la /home/user'')
\end{itemize}

\paragraph{Testing Results}{}
Table~\ref{tab:eval-classifier-results} summarizes the classification accuracy
and average per-query latency for each model:

\begin{table}[h]
  \centering
  \caption{Binary classification results on \texttt{classifier-100}.}
  \begin{tabular}{lcc}
    \toprule
    \textbf{Model} & \textbf{Accuracy} & \textbf{Avg Latency (s)} \\
    \midrule
    ibm/granite3.1-moe:1b & 53.0\% & 0.278 \\
    ibm/granite3.1-moe:3b & 60.0\% & 0.101 \\
    ibm/granite3.1-dense:2b & 83.0\% & 0.377 \\
    qwen3:0.6b & 79.0\% & 1.183 \\
    qwen3:1.7b & 88.0\% & 2.871 \\
    qwen2.5:0.5b & 50.0\% & 0.180 \\
    qwen2.5:1.5b & 78.0\% & 0.150 \\
    gemma3:1b & 65.0\% & 0.233 \\
    \bottomrule
  \end{tabular}\label{tab:eval-classifier-results}
\end{table}

\paragraph{Key Findings and Limitations}
Evaluation and testing of this component revealed critical limitations that
prevented deployment:

\begin{enumerate}
  \item \textbf{Accuracy Ceiling:} Even the best-performing model
    (qwen3:1.7b) reached only 88\% accuracy on a balanced dataset, while several
    smaller models remained near chance-level performance (50--65\%).
  \item \textbf{Latency vs.\ Accuracy Trade-off:} Higher accuracy required
    larger models with substantially higher latency (up to 2.9s per query),
    which is too slow for interactive shell use.
  \item \textbf{Inconsistent Behavior on Ambiguous Inputs:} Misclassifications
    were frequent on short or context-dependent prompts, which are common in
    CLI workflows.
\end{enumerate}

\paragraph{Conclusion on Component 1}
This component remains a \textbf{proof-of-concept} and was not
pursued further. The fundamental limitation is that small local
\glspl{llm} cannot reliably classify ambiguous inputs in real-time
without either: (a) unacceptable latency through multiple calls, or
(b) extensive hardcoding of program paths that reduce
generalizability, (c) the shell wrapper itself becomes a bottleneck
and maintenance nightmare. The project focus was therefore shifted to
the \gls{rag} component, which proved more practical and effective.

\subsection{Component 2: Document Retriever}

\paragraph{Evaluation Approach}
The retriever is evaluated on its ability to return relevant documents within
top-k results using Precision@k and MRR, with three automated test variants:
\begin{enumerate}
  \item Document-level retrieval (test1)
  \item Document retrieval with passage-level re-ranking (test2)
      \textit{(retrieving documents first, then re-ranking passages within them, evaluating by passage matches)}
  \item Passage-scope retrieval (test3)
\end{enumerate}
Test1 checks whether the correct document appears in the top-k results. Test2
first retrieves top-k documents and then re-ranks passages within them
(top\_k\_extended=10) to locate the expected section. Test3 indexes and retrieves
passages directly (passage\_size=20 lines), evaluating whether the correct
document is found via passage matches.

\paragraph{Testing Collections}
\begin{itemize}
  \item \texttt{sample-10}: 14 documents, 250 queries
  \item \texttt{sample-100}: 101 documents, 300 queries
\end{itemize}

\paragraph{Testing Results}{}
Tables~\ref{tab:eval-retriever-sample10-doc}--\ref{tab:eval-retriever-sample100-passage}
summarize document-level retrieval (test1), passage-level re-ranking (test2), and
passage-level retrieval (test3). All tests used top\_k=5; test2 uses
top\_k\_extended=10.

\begin{table}[h]
  \centering
  \caption{Document-level retrieval on \texttt{sample-10} (test1).}
  \begin{tabular}{lccccc}
    \toprule
    \textbf{Embedding Model} & \textbf{Top-1} & \textbf{Top-5} & \textbf{MRR} & \textbf{Avg Latency (s)} & \textbf{Index Build (s)} \\
    \midrule
    ibm/granite-embedding:30m & 83.6\% & 95.2\% & 0.881 & 0.071 & 0.348 \\
    ibm/granite-embedding:125m & 85.2\% & 92.4\% & 0.880 & 0.075 & 16.603 \\
    ibm/granite-embedding:278m & 84.8\% & 87.6\% & 0.861 & 0.099 & 2.292 \\
    qwen3-embedding:0.6b & 86.8\% & 100.0\% & 0.916 & 0.074 & 17.503 \\
    \bottomrule
  \end{tabular}\label{tab:eval-retriever-sample10-doc}
\end{table}

\begin{table}[h]
  \centering
  \caption{Document retrieval with passage re-ranking on \texttt{sample-10} (test2).}
  \begin{tabular}{lccccc}
    \toprule
    \textbf{Embedding Model} & \textbf{Top-1} & \textbf{Top-5} & \textbf{MRR} & \textbf{Avg Latency (s)} & \textbf{Index Build (s)} \\
    \midrule
    ibm/granite-embedding:30m & 48.4\% & 87.2\% & 0.628 & 0.817 & 1.138 \\
    ibm/granite-embedding:125m & 39.2\% & 87.6\% & 0.557 & 1.264 & 1.044 \\
    ibm/granite-embedding:278m & 46.4\% & 81.6\% & 0.594 & 1.514 & 1.900 \\
    qwen3-embedding:0.6b & 55.2\% & 94.0\% & 0.685 & 3.777 & 10.322 \\
    \bottomrule
  \end{tabular}\label{tab:eval-retriever-sample10-rerank}
\end{table}

\begin{table}[h]
  \centering
  \caption{Passage-level retrieval on \texttt{sample-10} (test3).}
  \begin{tabular}{lccccc}
    \toprule
    \textbf{Embedding Model} & \textbf{Top-1} & \textbf{Top-5} & \textbf{MRR} & \textbf{Avg Latency (s)} & \textbf{Index Build (s)} \\
    \midrule
    ibm/granite-embedding:30m & 83.6\% & 95.2\% & 0.881 & 0.028 & 1.075 \\
    ibm/granite-embedding:125m & 85.2\% & 92.4\% & 0.880 & 0.032 & 1.002 \\
    ibm/granite-embedding:278m & 84.8\% & 87.6\% & 0.861 & 0.099 & 2.134 \\
    qwen3-embedding:0.6b & 86.8\% & 100.0\% & 0.916 & 0.074 & 10.684 \\
    \bottomrule
  \end{tabular}\label{tab:eval-retriever-sample10-passage}
\end{table}

\begin{table}[h]
  \centering
  \caption{Document-level retrieval on \texttt{sample-100} (test1).}
  \begin{tabular}{lccccc}
    \toprule
    \textbf{Embedding Model} & \textbf{Top-1} & \textbf{Top-5} & \textbf{MRR} & \textbf{Avg Latency (s)} & \textbf{Index Build (s)} \\
    \midrule
    ibm/granite-embedding:30m & 62.7\% & 86.3\% & 0.724 & 0.043 & 2.270 \\
    ibm/granite-embedding:125m & 69.7\% & 88.7\% & 0.768 & 0.036 & 1.706 \\
    ibm/granite-embedding:278m & 65.7\% & 84.0\% & 0.734 & 0.098 & 1.652 \\
    qwen3-embedding:0.6b & 72.7\% & 92.0\% & 0.806 & 0.074 & 128.222 \\
    \bottomrule
  \end{tabular}\label{tab:eval-retriever-sample100-doc}
\end{table}

\begin{table}[h]
  \centering
  \caption{Document retrieval with passage re-ranking on \texttt{sample-100} (test2).}
  \begin{tabular}{lccccc}
    \toprule
    \textbf{Embedding Model} & \textbf{Top-1} & \textbf{Top-5} & \textbf{MRR} & \textbf{Avg Latency (s)} & \textbf{Index Build (s)} \\
    \midrule
    ibm/granite-embedding:30m & 42.7\% & 72.7\% & 0.530 & 0.595 & 2.132 \\
    ibm/granite-embedding:125m & 43.0\% & 77.0\% & 0.539 & 0.802 & 2.451 \\
    ibm/granite-embedding:278m & 45.7\% & 72.7\% & 0.549 & 1.192 & 3.219 \\
    qwen3-embedding:0.6b & 45.7\% & 79.7\% & 0.571 & 3.745 & 128.563 \\
    \bottomrule
  \end{tabular}\label{tab:eval-retriever-sample100-rerank}
\end{table}

\begin{table}[h]
  \centering
  \caption{Passage-level retrieval on \texttt{sample-100} (test3).}
  \begin{tabular}{lccccc}
    \toprule
    \textbf{Embedding Model} & \textbf{Top-1} & \textbf{Top-5} & \textbf{MRR} & \textbf{Avg Latency (s)} & \textbf{Index Build (s)} \\
    \midrule
    ibm/granite-embedding:30m & 62.7\% & 86.3\% & 0.724 & 0.037 & 2.200 \\
    ibm/granite-embedding:125m & 69.7\% & 88.7\% & 0.768 & 0.036 & 1.783 \\
    ibm/granite-embedding:278m & 65.7\% & 84.0\% & 0.734 & 0.101 & 1.705 \\
    qwen3-embedding:0.6b & 72.7\% & 92.0\% & 0.806 & 0.072 & 134.423 \\
    \bottomrule
  \end{tabular}\label{tab:eval-retriever-sample100-passage}
\end{table}

\paragraph{Key Findings}
The document retriever component proved robust and practical. On
\texttt{sample-10}, document-level Top-1 accuracy ranged from 83.6\% to 86.8\%
with Top-5 accuracy up to 100\%. On \texttt{sample-100}, Top-1 ranged from
62.7\% to 72.7\% with Top-5 up to 92.0\%. The qwen3-embedding:0.6b model
provided the best precision, but its index build time was substantially higher
on the larger corpus (128--134s vs.\ 1.6--2.3s for Granite embeddings).
% TODO: test1-3 comparison, context size

\subsection{Component 3: Information Summarizer}

\paragraph{Evaluation Approach}
The summarizer is evaluated on answer quality and relevance using
queries with reference answers.
% TODO: more details on evaluation metrics

\paragraph{Test Collection}
\texttt{summarize-30}: 30 queries with reference answers, derived from user
queries, from \texttt{sample-10} and \texttt{sample-100} document collections.

\paragraph{Testing Results}{}
Table~\ref{tab:eval-1-results} presents results from evaluation for each
collection:

\begin{table}[h]
  \centering
  \caption{Evaluation results on \ldots} % TODO
  \begin{tabular}{lccc}
    \toprule
    \textbf{Test} & \textbf{Metric} & \textbf{Result} & \textbf{Avg Latency} \\
    \midrule
    Test 1 & Metric 1 & Result 1 & Number \\
    \bottomrule
  \end{tabular}\label{tab:eval-1-results}
\end{table}
\ldots % TODO: see precision metrics for bin-cl

\paragraph{Key Findings}

\paragraph{Evaluation Approach}
The complete \gls{rag} system (document retrieval + information
summarization) is evaluated end-to-end to assess real-world
performance including document indexing time, query processing, and
user response latency under different hardware configurations and document
corpus sizes. The evaluation focuses on practical usability for \gls{cli}
assistance.

\paragraph{Testing Collection}{}
For end-to-end testing, we will reuse the \texttt{summarizer-150} collection of
queries, which will be executed on a set of devices with varying hardware

\paragraph{Testing Results}{}
Table~\ref{tab:eval-1-results} presents results from evaluation for each
collection:

\begin{table}[h]
  \centering
  \caption{Evaluation results on \ldots} % TODO
  \begin{tabular}{lccc}
    \toprule
    \textbf{Test} & \textbf{Metric} & \textbf{Result} & \textbf{Avg Latency} \\
    \midrule
    Test 1 & Metric 1 & Result 1 & Number \\
    \bottomrule
  \end{tabular}\label{tab:eval-1-results}
\end{table}
\ldots % TODO: see precision metrics for bin-cl

\paragraph{Key Findings}
% The complete \gls{rag} system demonstrates practical performance for
% \gls{cli} assistance:
% \begin{itemize}
%   \item Query latency: 250-700ms depending on model and document corpus size
%   \item Memory footprint: 2-4 GB for 100+ documents with embedding indices
%   \item Scalability: Tested with up to 200+ documents; performance
%     remains acceptable
%   \item User experience: The combination of fast retrieval + focused
%     summarization provides relevant, accurate answers to command-line
%     queries without the latency/accuracy issues of the classification component
% \end{itemize}

\paragraph{Limitations}
The \gls{rag} system is constrained to document-based assistance and cannot
handle task-specific guidance or dynamic system introspection. It works well
for: explaining commands, providing usage examples, and retrieving
documentation. It is less suitable for: system diagnostics, real-time
monitoring, or complex decision-making tasks that require agent-like
capabilities.

\subsection{Why Agent-Based Approaches Failed}

During exploration, agentic approaches (where the \gls{llm} could iteratively
reason, use tools, and take actions) were tested to handle more complex tasks.
These experiments revealed fundamental limitations of small local \glspl{llm}:

\begin{enumerate}
  \item \textbf{Reasoning Chain Breakdown:} Multi-step reasoning
    required to plan and execute agent actions frequently broke down
    in small models, with incorrect intermediate conclusions
    cascading into failures.
  \item \textbf{Tool Use Unreliability:} Even with explicit tool
    definitions, small models struggled with proper tool invocation
    syntax and parameter selection, requiring heavy prompt
    engineering and validation logic.
  \item \textbf{Hallucination Under Uncertainty:} When agents faced
    unexpected states or error conditions, models tended to
    hallucinate plausible-sounding responses rather than correctly
    diagnosing issues or requesting clarification.
  \item \textbf{Unacceptable Latency:} Achieving higher accuracy
    required extensive prompting, validation loops, and fallback
    mechanisms, resulting in total latencies of 5-15 seconds per query.
\end{enumerate}

\paragraph{Conclusion on Agent-Based Approaches}
Agent-based and classification-focused approaches require significantly larger
models (10B+ parameters) or external \gls{llm} services to function reliably.
For local, privacy-preserving assistance on consumer hardware, the program paths
should rather be programmatically constrained and suited for specific tasks
(e.g., retrieval + summarization) rather than open-ended reasoning or tool use.
